---
title: "Optimising Loan Eligibility: A Study of Machine Learning Models"
author: "Chandrika Kunchakuri, Satya Sai Gayathri Palacherla"
output:
  pdf_document: default
  html_document: default


---

**Background:**

Loan services play a crucial role in the modern world, serving as a primary source of profit for
financial institutions. They help students cover educational and living expenses and enable
individuals to acquire luxuries like homes and cars. However, determining whether an applicant
qualifies for a loan requires careful evaluation of various factors. The "Loan Prediction Problem
Dataset," sourced from Kaggle by Debdatta Chatterjee, offers a comprehensive set of metrics to 
analyze the determinants of loan approval. This study focuses on identifying the most effective
metric for loan approval using logistic regression and random forest models.

**Dataset Attributes:**
The data collection consists of 614 records and 13 attributes, which include:
Loan_ID: A unique identification number for each loan application.
Gender: The gender of the applicant (Male or Female).
Married: The marital status of the applicant (Yes or No).
Dependents: The number of dependents of the applicant (0, 1, 2, or 3+).
Education: The level of education of the applicant (Graduate or Not Graduate).
Self_Employed: Whether the applicant is self-employed or not (Yes or No).
ApplicantIncome: The monthly income of the applicant.
CoapplicantIncome: The monthly income of the co-applicant.
LoanAmount: The loan amount requested (in thousands).
Loan_Amount_Term: The term of the loan, in months.
Credit_History: A binary variable representing the applicant’s credit history (1 if they have a
credit history that meets the guidelines, 0 otherwise).
Property_Area: The category of the property area where the applicant resides (Urban, Semiurban,
or Rural).
Loan_Status: The final decision on the loan application (Y for approved, N for not approved).


```{r}
data<-read.csv("/Users/chandrikakunchakuri/Downloads/archive-2/train_u6lujuX_CVtuZ9i.csv",head=TRUE,stringsAsFactors = TRUE)
head(data)
str(data)
```
**3. Proposed Analysis**

**A. Data Preparation**

The below script performs the following steps to clean the data:

- Removes the first column.
- Replaces "3+" with 3 in the Dependents column.
- Checks for and removes rows with any missing values.
- Replaces missing values in the LoanAmount column with the mean value of that column.
- Checks the dimensions of the cleaned dataframe and re-checks for any remaining missing values.

```{r}
library(dplyr)
# DATA CLEANING:

data<-data[,-1]

data$Dependents <- ifelse(data$Dependents == "3+", 3, data$Dependents)


# Check for missing values
sum(is.na(data))

# Removing rows with missing values
data <- na.omit(data)
data$LoanAmount[is.na(data$LoanAmount)] <- mean(data$LoanAmount, na.rm = TRUE)
library(plyr)

dim(data)
sum(is.na(data))
```



```{r}

duplicated_rows <- duplicated(data)
duplicates <- data[duplicated_rows, ]
duplicates

data<-data[-470,]

dim(data)

head(data)
```




```{r}
str(data)
any(is.na(data))

```


Mostly, all the rows in the column Loan_Amount_Term is "360", so we are dropping it.

```{r}
data$Gender <- as.factor(data$Gender)
data$Married <- as.factor(data$Married)
data$Education <- as.factor(data$Education)
data$Self_Employed <- as.factor(data$Self_Employed)
data$Property_Area <- as.factor(data$Property_Area)
data$Loan_Status <- as.factor(data$Loan_Status)


table(data$Credit_History)
data$Credit_History <- as.factor(data$Credit_History)



table(data$Loan_Amount_Term) # Mostly all are 360 the column can be dropped 
data <- data[, -which(names(data) == "Loan_Amount_Term")]


```
```{r}

data$yes <- ifelse(data$Loan_Status=="Y",1,0)
data <- data[, -which(names(data) == "Loan_Status")]
head(data)
data$yes <- as.factor(data$yes)

data$ApplicantIncome <- as.numeric(data$ApplicantIncome)
str(data)
any(is.na(data))
```
**B. Exploratory Data Analysis (EDA)**

  - Univariate Analysis: Analyze the distribution of each variable individually (e.g., histograms for numeric variables, bar charts for categorical variables).
  - Bivariate Analysis: Examine the relationship between variables like “Loan_Status” and demographic features (e.g.,cross-tabulation, boxplots).
  - Multivariate Analysis: Investigate correlations between numeric variables (e.g., correlation matrix and heatmap).
  - Visualizations (e.g., scatter plots, boxplots) will be used to identify patterns and relationships.
  
```{r}
library(ggplot2)
# Define your data frame
library(ggplot2)
library(GGally)

ggpairs(data)

numeric_columns <- sapply(data, is.numeric)

ggpairs(data[, numeric_columns])
```
Overall, there are no very strong correlations between the variables, but some moderate and weak correlations exist. The most notable relationship is between ApplicantIncome and LoanAmount, indicating that applicant income plays a significant role in determining the loan amount. Other variables show weak or negligible correlations, suggesting that they might not be strong predictors of each other in the context of this dataset.

```{r}
# Add color based on a categorical variable (Credit_History)
ggpairs(data[, numeric_columns], aes(color = data$Credit_History))

```

Checking for the categorical values in the given Dataset.
```{r}
categorical_vars <- sapply(data, function(x) is.factor(x) || is.character(x))
categorical_vars
```


```{r}

library(ggplot2)
ggplot(data=data, aes(x=LoanAmount, fill=Education)) +
  geom_density() +
  facet_grid(Education~.)
```

Boxplot for the categorical values.

```{r}
par(mfrow=c(2,3))
counts <- table(data$yes, data$Gender)
barplot(counts, main="Loan Status by Gender",
        xlab="Gender", col=c("brown","lightblue"),
        legend = rownames(counts))
counts2 <- table(data$yes, data$Education)
barplot(counts2, main="Loan Status by Education",
        xlab="Education", col=c("brown","lightblue"),
        legend = rownames(counts2))
counts3 <- table(data$yes, data$Married)
barplot(counts3, main="Loan Status by Married",
        xlab="Married", col=c("brown","lightblue"),
        legend = rownames(counts3))
counts4 <- table(data$yes, data$Self_Employed)
barplot(counts4, main="Loan Status by Self Employed",
        xlab="Self_Employed", col=c("brown","lightblue"),
        legend = rownames(counts4))
counts5 <- table(data$yes, data$Property_Area)
barplot(counts5, main="Loan Status by Property_Area",
        xlab="Property_Area", col=c("brown","lightblue"),
        legend = rownames(counts5))
counts6 <- table(data$yes, data$Credit_History)
barplot(counts6, main="Loan Status by Credit_History",
        xlab="Credit_History", col=c("brown","lightblue"),
        legend = rownames(counts5))
```
```{r}
str(data)
any(is.na(data))
```
```{r}
boxplot(data$ApplicantIncome ~ data$yes,
        data = data,
        xlab = "Applicant income",
        ylab = "Loan Status",
        main = "Box plot applicant income vs Loan Status")

boxplot(data$CoapplicantIncome ~ data$yes,
        data = data,
        xlab = "Applicant income",
        ylab = "Loan Status",
        main = "Box plot Co applicant income vs Loan Status")

boxplot(data$LoanAmount ~ data$yes,
        data = data,
        xlab = "Loan Amount",
        ylab = "Loan Status",
        main = "Box plot Loan Amount vs Loan Status")

```

```{r}
library(corrplot)
numeric_data <- data[sapply(data, is.numeric)]
cor_matrix <- cor(numeric_data)

# Create a correlation plot
corrplot(cor_matrix, method = "color", order = "hclust",tl.col = "black", tl.srt = 45, addCoef.col = "black")


```
  
  
```{r}
#set.seed(123)  
#train_index <- sample(1:nrow(data), 0.7 * nrow(data))  # 70% for training
train_index=0.7 * nrow(data)
train_data <- data[1:train_index, ]
test_data <- data[train_index+1:ncol(data), ]
```

```{r}
#Logistic Regression

logistic_test<- glm (yes ~ ., data = train_data, family = binomial)

prediction_train <- predict(logistic_test, newdata = train_data[,c(1,2,3,4,5,6,7,8,9,10)] , type = "response")
prediction_train <- ifelse(prediction_train > 0.5,1,0)
#prediction_train

confusion_matrix_train <- table(train_data$yes, prediction_train )
confusion_matrix_train

sensitivity_train <- confusion_matrix_train[2, 2] / sum(confusion_matrix_train[2, ])
cat("Sensitivity for training data:", sensitivity_train, "\n")

# Calculate Specificity (True Negative Rate)
specificity_train <- confusion_matrix_train[1, 1] / sum(confusion_matrix_train[1, ])
cat("Specificity for training data:",specificity_train, "\n")

prediction_test <- predict(logistic_test, newdata = test_data[,c(1,2,3,4,5,6,7,8,9,10)] , type = "response")
prediction_test <- ifelse(prediction_test > 0.5,1,0)
#prediction_test

confusion_matrix_test <- table(test_data$yes, prediction_test)
confusion_matrix_test

sensitivity_test <- confusion_matrix_test[2, 2] / sum(confusion_matrix_test[2, ])
cat("Sensitivity for test data:", sensitivity_test, "\n")

# Calculate Specificity (True Negative Rate)
specificity_test <- confusion_matrix_test[1, 1] / sum(confusion_matrix_test[1, ])
cat("Specificity for test data:",specificity_test, "\n")


summary(logistic_test)

```

```{r}
odds_ratio_credit_history <- exp(coef(logistic_test)['Credit_History1'])
odds_ratio_credit_history

odds_ratio_ApplicantIncome <- exp(coef(logistic_test)['ApplicantIncome'])
odds_ratio_ApplicantIncome

odds_ratio_CoapplicantIncome <- exp(coef(logistic_test)['CoapplicantIncome'])
odds_ratio_CoapplicantIncome

odds_ratio_LoanAmount <- exp(coef(logistic_test)['LoanAmount'])
odds_ratio_LoanAmount


odds_ratio_Property_Area <- exp(coef(logistic_test)['Property_AreaSemiurban'])
odds_ratio_Property_Area

```

```{r}
# backward selection
bsel<-step(logistic_test,trace=0) 
formula(bsel)
summary(bsel)
```

```{r}
prediction_train <- predict(bsel, newdata = train_data[,c(1,2,3,4,5,6,7,8,9,10)] , type = "response")
prediction_train_binary <- ifelse(prediction_train > 0.5,1,0)
#prediction_train

confusion_matrix_train <- table(train_data$yes, prediction_train_binary )
confusion_matrix_train

sensitivity_train <- confusion_matrix_train[2, 2] / sum(confusion_matrix_train[2, ])
cat("Sensitivity for training data:", sensitivity_train, "\n")

# Calculate Specificity (True Negative Rate)
specificity_train <- confusion_matrix_train[1, 1] / sum(confusion_matrix_train[1, ])
cat("Specificity for training data:",specificity_train, "\n")

prediction_test <- predict(logistic_test, newdata = test_data[,c(1,2,3,4,5,6,7,8,9,10)] , type = "response")
prediction_test_binary <- ifelse(prediction_test > 0.5,1,0)
#prediction_test

confusion_matrix_test <- table(test_data$yes, prediction_test_binary)
confusion_matrix_test

sensitivity_test <- confusion_matrix_test[2, 2] / sum(confusion_matrix_test[2, ])
cat("Sensitivity for test data:", sensitivity_test, "\n")

# Calculate Specificity (True Negative Rate)
specificity_test <- confusion_matrix_test[1, 1] / sum(confusion_matrix_test[1, ])
cat("Specificity for test data:",specificity_test, "\n")
```
```{r}
# ROC curve for backward selection 
#install.packages("pROC")
library(pROC)


train_data$yes <- as.numeric(train_data$yes)
prediction_train<-as.numeric(prediction_train)
test_data$yes <- as.numeric(test_data$yes)
prediction_test<-as.numeric(prediction_test)

roc_curve_train <- roc(train_data$yes, prediction_train )
roc_curve_test <- roc(test_data$yes, prediction_test )

# Plotting ROC curve
plot(roc_curve_train, main = "ROC Curve Train", col = "blue")
#plot(roc_curve_test, main = "ROC Curve Test ", col = "red")
legend("bottomright", legend = paste("AUC =", round(auc(roc_curve_train), 2)), col = "blue", lty = 1, cex = 0.8)
#legend("bottomright", legend = paste("AUC =", round(auc(roc_curve_test), 2)), col = "blue", lty = 1, cex = 0.8)
library(pROC)

auc_value_train <- auc(train_data$yes,prediction_train)
cat("AUC value training data",auc_value_train)

#auc_value_test <- auc(test_data$yes,prediction_test)
#cat("AUC value test data",auc_value_test)
```
```{r}
odds_ratio_credit_history <- exp(coef(bsel)['Credit_History1'])
odds_ratio_credit_history


odds_ratio_LoanAmount <- exp(coef(bsel)['LoanAmount'])
odds_ratio_LoanAmount


odds_ratio_Property_Area <- exp(coef(bsel)['Property_AreaSemiurban'])
odds_ratio_Property_Area
```
```{r}
library(randomForest)

myyes_train<-as.factor(ifelse(train_data$yes==1,0,1))
myyes_test<-as.factor(ifelse(test_data$yes==1,0,1))


rf_model1 <- randomForest(myyes_train ~ ., data = train_data[,-11])

rf_model2 <- randomForest(myyes_train ~ Gender + Education + LoanAmount 
                          + Credit_History + Property_Area, 
                          data = train_data[,-11], ntree = 100)
rf_model3 <- randomForest(myyes_train ~ ApplicantIncome + CoapplicantIncome  
                          + LoanAmount + Credit_History + Property_Area 
                          ,data=train_data[,-11])


importance <- importance(rf_model1)
importance

rf_model3
```
```{r}
threshold1 <- function(predict, response) {
perf <- ROCR::performance(ROCR::prediction(predict, response),
"sens", "spec")
df <- data.frame(cut = perf@alpha.values[[1]], sens = perf@x.values[[1]],
spec = perf@y.values[[1]])
df[which.max(df$sens + df$spec), "cut"]
}
```

```{r}
predictions_train <- predict(rf_model1, newdata = train_data,type="prob")
predictions_train_binary <- ifelse(predictions_train[,2] > 0.5,1,0)
cm_train <- table(myyes_train,predictions_train_binary)
cm_train
sensitivity_train <- cm_train[2, 2] / sum(cm_train[2, ])
cat("Sensitivity for train data:", sensitivity_train, "\n")
specificity_train <- cm_train[1, 1] / sum(cm_train[1, ])
cat("Specificity for train data:",specificity_train, "\n")



predictions_test <- predict(rf_model1, newdata = test_data,type="prob")
predictions_test_binary <- ifelse(predictions_test[,2] > 0.5,1,0)
cm_test <- table(myyes_test, predictions_test_binary)
cm_test
sensitivity_test <- cm_test[2, 2] / sum(cm_test[2, ])
cat("Sensitivity for test data:", sensitivity_test , "\n")
specificity_test <- cm_test[1, 1] / sum(cm_test[1, ])
cat("Specificity for test data:",specificity_test, "\n")



pred=predict(rf_model1,type = "prob")
library(ROCR)
perf = prediction(pred[,2], myyes_train) #prob of predicting yes, target
# 1. True Positive and Negative Rate
roc1 = performance(perf,measure = "tpr",x.measure ="fpr")
# 2. Plot the ROC curve
plot(roc1,main="ROC Curve for Random Forest - Model 1",col=2,lwd=2,colorize=T)
abline(a=0,b=1,lwd=2,lty=2,col="gray")
#3. AUC
auc1 <- performance(perf, measure = "auc")
auc_ROCR <- auc1@y.values[[1]]
print(paste("AUC: ",round(auc_ROCR,4)))

```


```{r}
predictions_train <- predict(rf_model2, newdata = train_data,type="prob")
predictions_train_binary <- ifelse(predictions_train[,2] > 0.5,1,0)
cm_train <- table(myyes_train,predictions_train_binary)
cm_train
sensitivity_train <- cm_train[2, 2] / sum(cm_train[2, ])
cat("Sensitivity for train data:", sensitivity_train, "\n")
specificity_train <- cm_train[1, 1] / sum(cm_train[1, ])
cat("Specificity for train data:",specificity_train, "\n")



predictions_test <- predict(rf_model2, newdata = test_data,type="prob")
predictions_test_binary <- ifelse(predictions_test[,2] > 0.5,1,0)
cm_test <- table(myyes_test, predictions_test_binary)
cm_test
sensitivity_test <- cm_test[2, 2] / sum(cm_test[2, ])
cat("Sensitivity for test data:", sensitivity_test , "\n")
specificity_test <- cm_test[1, 1] / sum(cm_test[1, ])
cat("Specificity for test data:",specificity_test, "\n")



pred=predict(rf_model2,type = "prob")
library(ROCR)
perf = prediction(pred[,2], myyes_train) #prob of predicting yes, target
# 1. True Positive and Negative Rate
roc1 = performance(perf,measure = "tpr",x.measure ="fpr")
# 2. Plot the ROC curve
plot(roc1,main="ROC Curve for Random Forest - Model 2 ",col=2,lwd=2,colorize=T)
abline(a=0,b=1,lwd=2,lty=2,col="gray")
#3. AUC
auc1 <- performance(perf, measure = "auc")
auc_ROCR <- auc1@y.values[[1]]
print(paste("AUC: ",round(auc_ROCR,4)))


```

```{r}
predictions_train <- predict(rf_model3, newdata = train_data,type="prob")
predictions_train_binary <- ifelse(predictions_train[,2] > 0.5,1,0)
cm_train <- table(myyes_train,predictions_train_binary)
cm_train
sensitivity_train <- cm_train[2, 2] / sum(cm_train[2, ])
cat("Sensitivity for train data:", sensitivity_train, "\n")
specificity_train <- cm_train[1, 1] / sum(cm_train[1, ])
cat("Specificity for train data:",specificity_train, "\n")



predictions_test <- predict(rf_model3, newdata = test_data,type="prob")
predictions_test_binary <- ifelse(predictions_test[,2] > 0.5,1,0)
cm_test <- table(myyes_test, predictions_test_binary)
cm_test
sensitivity_test <- cm_test[2, 2] / sum(cm_test[2, ])
cat("Sensitivity for test data:", sensitivity_test , "\n")
specificity_test <- cm_test[1, 1] / sum(cm_test[1, ])
cat("Specificity for test data:",specificity_test, "\n")



pred=predict(rf_model3,type = "prob")
library(ROCR)
perf = prediction(pred[,2], myyes_train) #prob of predicting yes, target
# 1. True Positive and Negative Rate
roc1 = performance(perf,measure = "tpr",x.measure ="fpr")
# 2. Plot the ROC curve
plot(roc1,main="ROC Curve for Random Forest- model 3",col=2,lwd=2,colorize=T)
abline(a=0,b=1,lwd=2,lty=2,col="gray")
#3. AUC
auc1 <- performance(perf, measure = "auc")
auc_ROCR <- auc1@y.values[[1]]
print(paste("AUC: ",round(auc_ROCR,4)))

```


```{r}
# Load necessary libraries
library(glmnet)
library(caret)

# Load the dataset
data <- read.csv("Loan_Train.csv")

# Data Preprocessing
# Drop Loan_ID as it's not relevant for prediction
data <- data[, -which(names(data) == "Loan_ID")]

# Handle missing values
data$Gender[is.na(data$Gender)] <- "Male"
data$Married[is.na(data$Married)] <- "Yes"
data$Dependents[is.na(data$Dependents)] <- "0"
data$Self_Employed[is.na(data$Self_Employed)] <- "No"
data$LoanAmount[is.na(data$LoanAmount)] <- median(data$LoanAmount, na.rm = TRUE)
data$Loan_Amount_Term[is.na(data$Loan_Amount_Term)] <- median(data$Loan_Amount_Term, na.rm = TRUE)
data$Credit_History[is.na(data$Credit_History)] <- 1

# Convert Loan_Status to binary
data$Loan_Status <- ifelse(data$Loan_Status == "Y", 1, 0)

# Convert categorical variables to factors
data$Gender <- as.factor(data$Gender)
data$Married <- as.factor(data$Married)
data$Dependents <- as.factor(data$Dependents)
data$Education <- as.factor(data$Education)
data$Self_Employed <- as.factor(data$Self_Employed)
data$Property_Area <- as.factor(data$Property_Area)

# Prepare data for glmnet
x <- model.matrix(Loan_Status ~ ., data)[, -1]  # Remove intercept column
y <- data$Loan_Status

# Split data into training and test sets
set.seed(123)
trainIndex <- createDataPartition(y, p = 0.8, list = FALSE)
x_train <- x[trainIndex, ]
y_train <- y[trainIndex]
x_test <- x[-trainIndex, ]
y_test <- y[-trainIndex]

# Fit Lasso Regression with cross-validation to find the best lambda
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1, family = "binomial")

# Print the best lambda (regularization parameter)
cat("Best lambda (regularization parameter):", lasso_cv$lambda.min, "\n")

# Fit the final Lasso model using the best lambda
lasso_model <- glmnet(x_train, y_train, alpha = 1, family = "binomial", lambda = lasso_cv$lambda.min)

# Print the regression coefficients for all variables
cat("Regression Coefficients at Best Lambda:\n")
print(coef(lasso_model))

# Identify selected features (non-zero coefficients)
selected_features <- rownames(coef(lasso_model))[which(coef(lasso_model) != 0)]
cat("Selected Features (non-zero coefficients):\n")
print(selected_features)

# Predict on the test data using the fitted model
predictions <- predict(lasso_model, newx = x_test, type = "response", s = lasso_cv$lambda.min)

# Convert probabilities to binary class labels (threshold = 0.5)
predicted_classes <- ifelse(predictions > 0.5, 1, 0)

# Evaluate the model's performance
conf_matrix <- confusionMatrix(as.factor(predicted_classes), as.factor(y_test))
cat("Confusion Matrix:\n")
print(conf_matrix)


# Calculate Mean Square Error (MSE)
mse <- mean((y_test - predictions)^2)
cat("Mean Square Error (MSE):", mse, "\n")


# Print accuracy
cat("Model Accuracy:", conf_matrix$overall['Accuracy'], "\n")
```
**Literature Review**
  **Referrence 1**: Credit Risk Assessment and Financial Decision Support Using Explainable Artificial Intelligence” published by MDPI. This research discusses the application of machine learning models, including logistic regression and random forest, for credit risk assessment, emphasizing the importance of model interpretability in financial decision-making. The study uses Shapley values for explainability, enhancing model transparency in credit risk management. We can find more information in the URL "https://www.mdpi.com/2227-9091/12/10/164".
  
  
  **Referrence 2**: A Comparison of Random Forest and Logistic Regression Model in Credit Scoring of Rural Households.” This study compares the effectiveness of random forest and logistic regression models in predicting creditworthiness for rural households, focusing on the interpretability of logistic regression and the predictive accuracy of random forest. For more information, we can visit the URL "https://www.researchgate.net/profile/Do-Nhung-7/publication/353326832_A_COMPARISON_OF_RANDOM_FOREST_AND_LOGISTIC_REGRESSION_MODEL_IN_CREDIT_SCORING_OF_RURAL_HOUSEHOLDS/links/60f47053fb568a7098bd11e2/A-COMPARISON-OF-RANDOM-FOREST-AND-LOGISTIC-REGRESSION-MODEL-IN-CREDIT-SCORING-OF-RURAL-HOUSEHOLDS.pdf?origin=publication_detail".

